A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data (Hadoop Or Not Hadoop)

### 2014-02-14 text1.tex
Version on 14.02.14

%Current practices: HPC systems
A common step of NGS data analysis consists of mapping short reads to a reference sequence and then finding the genetic variations specific to the sample. Most of the bioinformatic programs are written for the Linux operating system. Some of the most widespread tools \footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000} like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made up with distributed computing in mind; many others do not even have the native ability to use multiple cores on the same computer\footnote{Samtools  acquired this feature since v.0.1.19}.

@@

The most common approach to speed up NGS tools is to parallelise within a compute node using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node which usually does not exceed~16.  For the tools which do not support the OMP natively, i.e. Samtools, variant calling can be parallelised by creating a separate process  for  each chromosome, or using GNU Parallel~\cite{gnuparallel} Linux utility.
Of great importance is that a multi-core approach does not improve the performance of operations that are limited by disk or network throughput, motivating to split  the dataset and  use multi-node parallelisation. Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since all the threads synchronisation (or load balancing) has to be coded by a programmer and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.

Another common way to introduce parallelisation in Linux systems is to use Bash scripting e.g. involve already existing utilities and cluster tools, split the data into chunks and process them on the separate nodes, and merging the results afterwards. This kind of solution benefits from both MPI and OMP and provides good performance, but the developments requires substantial expertise in order to be efficient, and the process is tightly coupled to the local computational cluster and network architectures and in many cases cannot be re-used in other settings.

@@

%Manuscript focus: When is Hadoop currently an appealing alternative
%Introduce what we did in the manuscript.
In this manuscript we focus on the question {\it if} and {\it when} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis. Since Hadoop is written in Java, which is slower than the standard~HPC programming languages like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective. We use five datasets with short reads of different sizes, align them against the reference genome, and call the variances.
%From the algorithmic point of view we propose a modification of a code for a preprocessing stage for Crossbow in order to benefit fully from massively parallel nature of MR computations. 
For the classical HPC DNA-seq analysis programs we developed a set of bash scripts utilizing multiple nodes for short reads alignment and exploiting the network fully, thereby speeding up calculations.
The execution times for each dataset were collected and scaling relations were analysed to answer the question in focus.

@@

\subsection{Analysis pipelines}
We constructed two pipelines for identifying SNPs from short read data, one based on Hadoop and one on regular HPC with a batch system (hereafter referred to as HPC). Our experiments then consisted of running the pipelines for the selected input dataset and measuring the wall-clock run time for each pipeline stage. All experiments were repeated several times and the results averaged to obtain a data point for the particular combination of data size and computational platform. Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (i.e. GATK~\cite{gatk}), we decided to create the  analysis pipeline as simple as possible to be able to pass the same stages on Hadoop and HPC:

@@

The equivalent Hadoop-based pipeline was implemented with Crossbow. The input data and the indexed genome reference were copied to Hadoop's storage system~(HDFS) before starting the experiments. Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.  Unfortunately Crossbow's preprocessor is not written in MR manner, and thus cannot be run in a massively parallel way. Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments. To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation, though sacrificing some generality\footnote{we assume that the FASTQ data are BZIP archived, and delivered to the storage accessible by Hadoop.}. To illustrate the benefits of this modification: For our 30-Gbase test dataset (approximately 45 GB of uncompressed FASTQ data) the preprocessing time shrunk from 3.3 hours to just under 2 minutes when run on 112 cores Hadoop cluster. The scripts for the preprocessing stage are publicly available~\cite{code_repo}.

\subsection{Computational resources}
To run the HPC analysis pipeline we used a computational cluster at UPPMAX, having nodes equipped with 4 dual-core CPUs. Data and reference genomes were stored on a parallel shared storage system. The Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud manager. The cluster was set up with Cloudera Hadoop distribution version~2.0.0-cdh4.4.0~\cite{cloudera}. For details on computational resources, see Supplementary material.

@@

\subsection{Accuracy of pipelines}
Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp correspondingly) we should not expect them to deliver perfectly matching SNPs lists, but still we expect them to capture and correctly identify the mutation. We tested the correctness just for the smallest dataset~I. The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$ was successfully localized by both applications\cite{schneeberger}. [IS THIS REFERENCE CORRECT AND IN RIGHT PLACE?]

\subsection{Scalability of HPC and Hadoop approaches}
To demonstrate the scalability of  the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:hpctiming}).
As one can see the aligning process with the Bowtie scales linearly, but the SNP calling part is a definite bottleneck, showing almost no scaling at all.
For the given datasets (I--V) the timings for alignment against the corresponding genomes and SNP calling for  the HPC and Hadoop approaches were collected (Table~\ref{table:2}).

One of the attractive sides of using Hadoop is its almost linear scalability,~i.e. calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, regardless the scaling nature of the underlying program (Samtools, Bowtie etc.). 
Figure~\ref{fig:fig1} shows the calculation time as a function of the dataset size for $p=56$ cores Hadoop cluster. Dataset V was excluded since it is for {\it H.sapiens}, which has more than 20 times larger genome than~{\it A.thaliana}.
To stress the linear nature of scaling, both sets of points were fit to linear polynomial using the least squares method.

\begin{table}[b]
\small
\caption{Calculation time for Dataset I executed on different  number of cores on a node of  HPC1 cluster. }
\begin{center}
\begin{tabular}{|c|c|c|c|}
$N$ cores	&\multicolumn{3}{c|}{timing, minutes}\\
\hline
	& aligning 	&	SNP calling\tablefootnote{We do not consider here special tricks how to parallelize the Samtools analysis by chromosome, as could be done, as exemplified here \url{http://www.biostars.org/p/48781/
}}	&	total \\
\hline
1	&	38	&	37		&	75 \\
2	&	19	&	36		&	55\\
4	&	10	&	36		&	46\\
8	&	5	&	36		&	41\\
\end{tabular}
\end{center}
\label{table:hpctiming}
\normalsize
\end{table}

\begin{table}[htdp]
\caption{Timings  (in minutes) for HPC and  Hadoop deployments for different dataset sizes.    Dataset is shown in brackets by roman numerals, ``f.r.'' stands for ``forward reads''.}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}

data, Gbases&	1.4&	3.5&	7.0 	&15.0 	&30.0 	&100.0 	&250 	\\
			&	(I)			& (II, f.r.)	&	(II)		&(III, f.r.) &(III)&(IV) &(V)\\
\hline
Hadoop, 56 cores&--&	--	&39		&62	&108	&250&1125\\
\hline
%Hadoop, 56 cores, Shared~TM&	25	&	24	&40		&	70&	150	&410&\\
%\hline
%Hadoop, 56 cores, ``bare-metal''  &	36	&	41	&53  	&63	&	92& 380	&\\
%\hline
HPC1, 8 cores&	41&	96	&157	&307	&596	&1490&--\\
%\hline
%HPC2, 8 cores&72	&	&205	&&413&1765&\\
%\hline
%$F_{Hadoop}/F_{HPC1}$, Shared TM &	4.27		&1.75	&1.78	&	1.60	&	1.75	&	1.93&\\
%\hline
%$F_{Hadoop}/F_{HPC1}$, SSH TM &&		&1.74	&1.41	&1.28	&1.17	&\\
%\hline
%$F_{Hadoop}/F_{HPC2}$, ``bare-metal'' &&&&&&&\\

%data, 	& \multicolumn{2}{c|}{Hadoop, 56 cores}		  	&  HPC			&	\multicolumn{2}{c|}{$F_{Hadoop}/F_{HPC}$}\\
%Gbases 	&	 SSH~TM,	&	Shared~TM,	&	(8 cores), 	&	Shared TM	& 	SSH TM\\
%&  minutes& minutes&  minutes& &\\

\end{tabular}
\end{center}
\label{table:2}
\end{table}%

\begin{figure}
	\input{./figures/fig1/fig1.tex}
	\caption{Calculation time depending on the dataset size for fixed size~($p=56$ cores) of Hadoop cluster on the private Cloud. The linear fit based on least squares method are provided, revealing almost linear scaling.}
	\label{fig:fig1}
\end{figure}

\subsection{Comparing Hadoop and HPC runtime efficiency for different dataset sizes}
Hadoop initially was designed to digest huge datasets\cite{hadoop,lin2010}. One can propose then that the larger the dataset is, the more suitable Hadoop becomes compare to the HPC approach. In order to compare the ''suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on different amount of cores, we constructed the following function:
$$F=T_{p}\times p,$$ 
where~$T_{p}$ is the calculation time on~$p$ cores.
Using the data from Table~\ref{table:2},  we plot the ratio $F_{Hadoop}/F_{HPC}$, keeping in mind that the closer the ratio to the unity, the closer Hadoop's efficiency is to that of the HPC approach.

\begin{figure}
	\input{./figures/fig2/fig2.tex}
	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases.  Calculations were carried out for~$p=56$ and $p=8$ cores for Hadoop and HPC1 correspondingly.
The points are fit in a line using least-squares method, what makes  possible to make some careful predictions for the case of the {\it infinite} dataset size.}
	\label{fig:fig2}
\end{figure}

The curve  in~Figure~\ref{fig:fig2} is plotted for~$p=56$ cores Hadoop cluster and an HPC node ($p=8$), and displays the ratio~$F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size. The extrapolation to the zero of the~$X$-axis tells the ratio for the hypothetical infinite dataset. As one can see, the Hadoop approach becomes more  and more effective compared to the HPC scenario as the data is larger. With  the linear extrapolation  the ratio reads~$1.13\pm0.01$, meaning that Hadoop running even  in the virtualized environment of a private cloud, is competitive with  the HPC approach which is being run on ''bare metal''  for the datasets greater than $100$\,Gbases (IV), which is a usual size for human genome sequencing giving the average sequencing depth of  about~$30x$.

@@

Due to the trivially parallelisation nature of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computational resources, i.e. split the initial data into chunks to process them independently.
Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost simultaneously have to be sent over the network and thereby increasing the communication costs,~$T_{comm}$.

There are several program packages for short read alignment designed with MPI support~\cite{pmap, gnumap}. Authors report almost linear scaling up to 32 nodes for pair-ended reads\footnote{\url{http://bmi.osu.edu/hpc/slides/Bozdag10-HiCOMB.pdf}, \url{http://dna.cs.byu.edu/gnumap/HICOMB_Presentation.pdf}}. However,  i.e. Pmap package functions poorly on UPPMAX cluster for  datasets larger than 20 Gbases, raising memory exceptions.
We implemented a highly optimized bash script making use of standard Unix utilities to use the  HPC cluster network as efficiently as possible~\cite{repo},  and compared the network performance with the standard Hadoop HDFS approach. We separated the mapping time~$T_{mapping}$ and the  communication time~ $T_{comm}$, and plot their ratio~$T_{mapping}/T_{comm}$ as a function of a reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}). 
Such kind of measure is applicable to both HPC and Hadoop, however, the~$T_{comm}$ carries a different sense in both cases. For Hadoop the short reads in FASTQ format have to be preprocessed (involving nodes communication~$T_{comm}$) to be able to run in MR-fashion, while the data locality will be automatically achieved during the data  ingestion into the HDFS. 
We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelisation.
For the HPC approach, the~$T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual mapping happens, and the sending of the aligned SAM files back to the delivery location over the network.

@@

\textit{Hadoop results (filled circles)}: One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40. It is known that the Bowtie provides a linear scaling between the mapping time and the dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{bowtie}, and Figure~\ref{fig:fig1}. Since the ratio~$T_{mapping}/T_{comm}\approx const$, one can conclude that the~$T_{comm}\propto 1/N$, meaning that the more nodes are involved, the faster the communication is in the preprocessing stage.

>>>>

%Current practices: HPC systems
A common step of NGS data analysis consists of mapping short reads to a reference sequence and then finding the genetic variations specific to the sample. Most of the bioinformatic programs are written for the Linux operating system. Some of the most widespread tools\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000; from internal communications} like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made up with distributed computing in mind; many others do not even have the native ability to use multiple cores on the same computer\footnote{Samtools  acquired this feature since v.0.1.19}.

@@

The most common approach to speed up NGS tools is to parallelise within a compute node using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node which usually does not exceed~16.  For the tools which do not support the OMP natively, i.e. Samtools, variant calling can be parallelised by creating a separate process  for  each chromosome, or using GNU Parallel~\cite{gnuparallel} Linux utility.
Of great importance is that a multi-core approach does not improve the performance of operations that are limited by disk or network throughputs, motivating to split  the dataset and  use multi-node parallelisation.

Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since all the threads synchronisation (or load balancing) has to be coded by a programmer and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.

Another common way to introduce parallelisation to NGS analysis pipelines in Linux systems is to use Bash scripting e.g. involve already existing utilities and cluster tools, split the data into chunks and process them on the separate nodes, and merging the results afterwards. This kind of solution benefits from both MPI-like  and OMP and provides good performance, but the development requires substantial expertise in order to be efficient, and the process is tightly coupled to the local computational cluster and network architectures and in many cases cannot be re-used in other settings.

@@

%Manuscript focus: When is Hadoop currently an appealing alternative
%Introduce what we did in the manuscript.
In this manuscript we focus on the question {\it if} and {\it when} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis. Since Hadoop is written in Java, which is slower than the standard~HPC programming languages like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective. We use five datasets with short reads of different sizes, align them against the reference genome, and call the variances.
For the existing Hadoop  software we propose modifications, in order to benefit fully from massively parallel nature of MR computations. 
For the classical HPC DNA-seq analysis programs we developed a set of Bash scripts utilizing multiple nodes for short reads alignment and exploiting the network fully, thereby speeding up calculations.
The execution times for each dataset were collected and scaling relations were analysed to answer the question in focus.

@@

\subsection{Analysis pipelines}
We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data, one based on Hadoop and one on regular HPC with a batch processing system (hereafter referred to as HPC). Our experiments then consisted of running the pipelines for the selected input dataset and measuring the wall-clock run time for each pipeline stage. All experiments were repeated several times and the results averaged to obtain a data point for the particular combination of data size and computational platform. Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (i.e. GATK~\cite{gatk}), we decided to create the  analysis pipeline as simple as possible to be able to pass the same stages on Hadoop and HPC:

@@

The equivalent Hadoop-based pipeline was implemented with Crossbow. The input data and the indexed genome reference were copied to Hadoop's storage system~(HDFS) before starting the experiments. Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.  Unfortunately Crossbow's preprocessor is not written in MR manner, and thus cannot be run in a massively parallel way. Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments. To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation.

\subsection{Computational resources}
To run the HPC analysis pipeline we used a computational cluster at UPPMAX, having nodes equipped with 8 dual-core CPUs. Data and reference genomes were stored on a parallel shared storage system\cite{gulo}. The Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud manager. The cluster was set up with Cloudera Hadoop distribution version~2.0.0-mr1-cdh4.5.0~\cite{cloudera}. For details on computational resources, see Supplementary material.

@@

\subsection{Modified preprocessing stage}
A native preprocessing stage in the Crossbow\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis} provides great flexibility in delivering the data to the Hadoop cluster.  Data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet\cite{crossbow}. The only way to introduce parallelisations in the native preprocessing stage~-- is to split the read files into smaller chunks and to generate a manifest file listing all these chunks. This, however, is lacking the massively parallel way of treating the data.

Being  pragmatic we assume that sequencing platforms are usually affiliated with computation facilities which provide both~--  storage and  computers,  
therefore considered the data already been downloaded to the storage Hadoop can access.
We rewrote the preprocessing stage scripts in a MR-manner, benefiting from its massively parallel nature.
The demand is that the FASTQ data are BZIP2 archived, and  accessible by Hadoop. For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were ingested to the HDFS.
The BZIP2 provides very good compression and is splittable, meaning that the archive can be expanded in a parallel manner\footnote{other options like splittable LZO are also possible}. Also the BZIP2 format is natively supported by Hadoop by enabling the respective codec, i.e. for the developer there is no difference in dealing with the FASTQ data or with its BZIP2 archive.

The Hadoop streaming library offers a possibility to write  Mapper and Reducer for Hadoop jobs in any  programming language. Our Python scripts efficiently process  short reads, produced by  Illumina sequencing platform of different versions, as well as the FASTQ files converted from the SRA format, NCBI Sequence Read Archive. 
The problem of lacking the unique FASTQ header standard was solved by the rewriting the header based on a SHA-1 hash function, and the reads mates were labeled with ``.1" and ``.2" suffixes. 
To ensure that the both forward and reverse read of the same pair will end up on the same Reducer, the secondary Mapper key-sorting mechanisms were involved.

In order to compare the native preprocessor and the proposed one, the data was put on the HDFS in beforehand, to eliminate possible network delays i.e. while downloading data from some Amazon cloud.
Table~\ref{table:preprocess} illustrates the benefits of our approach.
\begin{table}[htdp]
\small
\caption{Timings (in minutes) with the  Crossbow native read preprocessor and with derived approach for  datasets II, III, IV  on~$p=56$ cores Hadoop cluster, using the Crossbow, and with the BASH script on a single HPC node, $p=16$ cores.}
\begin{center}
\begin{tabular}{|l|c|c|c|c|}
dataset			&		II (7 Gbases )	& III (30 Gbases)	& IV (100 Gbases)\\
\hline
Crossbow native			&		$60.6\pm0.7$	& $299.0\pm2.3$	&	$673\pm1$	\\
Hadoop, this work			&		$7.0\pm0.0$	&	$20.7\pm0.4$&		$52.4\pm0.1$\\
BASH, this work			& 		$7.4\pm0.0$	&	$31.1\pm0.1$	&	$114.5\pm0.3$	\\
\end{tabular}
\end{center}
\label{table:preprocess}
\normalsize
\end{table}%
To complete the comparison the same preprocessing was performed with the  BASH scripting against the data located on the HPC storage. The script\footnote{\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}} ran on a single HPC node, utilizes  multiple cores, with the limiting factor being the  HDD IO performance, which effectively leaves with 5-7 cores of 16.

\subsection{Accuracy of pipelines}
Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp correspondingly) we should not expect them to deliver perfectly matching SNPs lists, but still we expect them to capture and correctly identify the mutation. We tested the correctness just for the smallest dataset~I. The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$\cite{schneeberger} was successfully localized by both applications.

\subsection{Scalability of HPC and Hadoop approaches}
To demonstrate the scalability of  the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}).
As one can see the aligning process with the Bowtie scales fairly good, but the SNP calling part is a bottleneck,  scaling worse than the Bowtie, and consuming progressively larger part of the total calculation time
%\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance. During its sorting phase the lack of RAM results in spilling to the HDD. For example for an half of  Dataset III, SNP calling takes approximately 40 minutes with 8 GB RAM, and 15 minutes with 48 GB RAM, when  the whole BAM file fits into RAM. For our tasks we used the default Samtool's settings.}.  
For the given datasets (I--V) the timings for alignment against the corresponding genomes and SNP calling for  the HPC and Hadoop approaches were collected (Table~\ref{table:4}).

One of the attractive sides of using Hadoop~MR is its almost linear scalability,~i.e. calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, regardless the scaling nature of the underlying program (Samtools, Bowtie etc.). 
Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for $p=56$ cores Hadoop cluster. Dataset V was excluded since it is for {\it H.sapiens}, which has more than 20 times larger genome than~{\it A.thaliana}.
To stress the linear nature of scaling, both sets of points were fit to linear polynomial using the least squares method.

\begin{table}[htb]
\small
\caption{Calculation time for Dataset I executed on different  number of cores on a node of  HPC cluster. }
\begin{center}
\begin{tabular}{|c|c|c|c|ccc|}
$N$ cores	&\multicolumn{3}{c|}{timing, minutes}&\multicolumn{3}{c|}{speed-ups} \\
\hline
	& mapping 	&	SNP calling\tablefootnote{We  used here special tricks how to parallelize the Samtools analysis by chromosome, as exemplified here \url{http://www.biostars.org/p/48781/
}}	&	total  &total & map& SNP call\\
\hline
1	&	22.75	$\pm$1.24&	26.1$\pm$0.7	&	48.9	&	 1.00	&	1.00	&	1.00\\
2	&	11.00$\pm$0.05	&	16.9$\pm$1.3	&	27.9	&	 0.88	&	1.03	&	0.77\\
4	&	5.91$\pm$0.04	&	11.7$\pm$0.5	&	17.6	&	 0.69	&	0.96	&	0.56\\
6	&	4.24$\pm$0.04	&	8.6$\pm$0.8	&	12.8	&	 0.64	&	0.89	&	0.51\\
8	&	3.44$\pm$0.03	&	8.6$\pm$0.9	&	12.1	&	 0.51	&	0.83	&	0.38\\
10	&	2.91$\pm$0.01	&	7.7$\pm$0.5	&	10.6	&	 0.46	&	0.78	&	0.34\\
12	&	2.64$\pm$0.04	&	7.5$\pm$0.5	&	10.1	&	 0.40	&	0.72	&	0.29\\
14	&	2.57$\pm$0.03	&	7.5$\pm$0.6	&	10.1	&	 0.35	&	0.63	&	0.25\\
16	&	2.88$\pm$0.06	&	7.4$\pm$0.4	&	10.2	&	 0.30	&	0.49	&	0.22\\
\end{tabular}
\end{center}
\label{table:3}
\normalsize
\end{table}

\begin{table}[htdp]
\small

\caption{Timings  (in minutes) for HPC and  Hadoop deployments for different dataset sizes.    Dataset is shown in brackets by roman numerals, ``f.r.'' stands for ``forward reads''. Big deviation for the HPC is due to Samtools BAM sorting, which is IO and memory intensive, thus strongly depend on the queuing at the HPC cluster.}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}

data, Gbases		&	1.4	&	3.5		&	7.0		&	15.0		&	30.0		&	100.0	&	250 	\\
				&	(I)	&	(II, f.r.)	&	(II)		&	(III, f.r.)	&	(III)		&	(IV)		&	(V)\\
\hline
Hadoop, 56 cores&--&	--	&39		&62	&108	&250&1125\\
Hadoop, 56 cores		&	$18\pm0	$	&	$18\pm0	$	&	$29\pm0$	&	$47\pm0	$	&	$89\pm0$	&	$238\pm1$		&	$1164\pm14$\\
HPC1, 8 cores&	41&	96	&157	&307	&596	&1490&--\\
HPC, 16 cores	&	$17\pm1$	&	$22\pm6$	&	$43\pm3$	&	$81\pm9$	&	$172\pm15$		&	$467\pm60$	& $>48$ hours\\

\end{tabular}
\end{center}
\label{table:4}
\normalsize
\end{table}%

\begin{figure}
	\input{./figures/fig1/fig1.tex}
	\caption{Calculation time depending on the  size for selected datasets for~$p=56$ cores Hadoop cluster on the private Cloud. The linear fit based on least squares method are provided, revealing  linear scaling. Datasets are marked with roman numerals.}
	\label{fig:fig1}
\end{figure}

\subsection{Comparing Hadoop and HPC runtime efficiency for different dataset sizes}
Hadoop initially was designed to digest huge datasets\cite{hadoop,lin2010}. One can propose then that the larger the dataset is, the more suitable Hadoop becomes compare to the HPC approach. In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on different amount of cores, we constructed the following function:
$$F=T_{p}\times p,$$ 
where~$T_{p}$ is the calculation time on~$p$ cores.
Using the data from Table~\ref{table:4},  we plot the ratio $F_{Hadoop}/F_{HPC}$, keeping in mind that the closer the ratio to the unity, the closer Hadoop's efficiency is to that of the HPC approach.
\begin{figure}
	\input{./figures/fig2/fig2.tex}
	\caption{The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal dataset size in Gigabases.  Calculations were carried out for~$p=56$ and $p=16$ cores for Hadoop and HPC correspondingly.
The points are fit in a curve using least-squares method, what makes  possible to make some careful predictions for the case of the {\it infinite} dataset size. Datasets are marked with roman numerals.}
	\label{fig:fig2}
\end{figure}
The curve  in~Figure~\ref{fig:fig2}, based on Table~\ref{table:4}, is plotted for~$p=56$ cores Hadoop cluster and an HPC node ($p=16$), and displays the ratio~$F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size. The extrapolation to the zero of the~$X$-axis tells the ratio for the hypothetical infinite dataset. As one can see, the Hadoop approach becomes more  and more effective compared to the HPC scenario as the data is larger. 
The parabolic extrapolation for  the ratio reads~$1.70\pm0.01$, meaning that Hadoop running even  in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with  the HPC approach which is being run on ``bare metal''  on the modern hardware for the datasets greater than $100$\,Gbases (dataset~IV), which is a usual size for human genome sequencing giving the average sequencing depth of  about~$30x$.

The simplest curve to fit the points nicely is a parabolic function, importantly not a linear one. Taking into account  a {\it linear} scaling for the Hadoop approach for those datasets, Figure~\ref{fig:fig1}, one can conclude that HPC approach scales worse than linearly\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e. being a constant; thus the ratio of two constants would result a constant, but we observe a dependecy.}. We suspect the main reason for such a behavior is the fact that for large datasets (starting from dataset~III in our case) the SNP calling routine with the Samtools becomes more memory and time demanding than the actual mapping with Bowtie.
If lacking  RAM, the Samtools spilling to HDD, creating multiple (up to hundreds) temporary files while sorting the BAM file, keeping the network load for NFS storage and  time expensive IO on a high rate.
The Table~\ref{table:5} shows the role of the amount of  allocated RAM on the Samtools sort timing, as a most time expensive,  for the dataset~IV, starting from 8GB up to 80GB, performing on a single HPC node. 
\begin{table}[htdp]
\small

\caption{Timings (in minutes) for the sort routine of the Samtools package for the dataset IV (58Gb BAM file). The measurements are done on a single HPC node with $p=16$ cores. Averaging is done over 5 independent simulations. The RAM amount is given for the all  16 cores. }
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
RAM, GB		&	8			&		16			&			32		& 		64				\\
timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
\end{tabular}
\end{center}
\label{table:5}
\normalsize
\end{table}%
A large uncertainty for  small amount of RAM we believe is due to other users' load  on the cluster network.
As the  result a large amount of RAM per node is needed in order to use Samtools effectively, while Hadoop accomplishes SNP calling  on much economic hardware on linear time  with just 12GB RAM per VM node.

@@

Due to the trivially parallelisation nature of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computational resources, i.e. split the initial data into chunks to process them independently.
Reducing the size of each data chunk reduces the aligner  job,~$T_{mapping}$, but at the same time, the more chunks almost simultaneously have to be sent over the network, potentially causing traffic jams, and therefore increasing the communication costs,~$T_{comm}$.

There are several program packages for short read alignment designed with MPI support~\cite{pmap, gnumap}. Authors report almost linear scaling up to 32 nodes for pair-ended reads\footnote{\url{http://bmi.osu.edu/hpc/slides/Bozdag10-HiCOMB.pdf}, \url{http://dna.cs.byu.edu/gnumap/HICOMB_Presentation.pdf}}. However,  i.e. Pmap package functions poorly on UPPMAX cluster for  datasets larger than 20 Gbases, raising memory exceptions.
We implemented a highly optimized Bash script making use of standard Unix utilities to use the  HPC cluster network as efficiently as possible~\cite{repo},  and compared the network performance with the standard Hadoop HDFS approach. We separated the mapping time~$T_{mapping}$ and the  communication time~ $T_{comm}$, and plot their ratio~$T_{mapping}/T_{comm}$ as a function of a reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}). 
Such kind of measure is applicable to both HPC and Hadoop, however, the~$T_{comm}$ carries a different sense in both cases. For Hadoop the short reads in FASTQ format have to be preprocessed (involving nodes communication~$T_{comm}$) to be able to run in MR-fashion, while the data locality will be automatically achieved during the data  ingestion into the HDFS. 
We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelisation.
For the HPC approach, the~$T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual mapping happens, and the sending of the aligned SAM\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}} files back to the delivery location over the network.

@@

\textit{Hadoop results (filled circles)}: One can see that the ratio~$T_{mapping}/T_{comm}$ reveals very weak dependency in a wide range of number of nodes~$N$: from 4 up to 40. It is known that the Bowtie provides almost  linear scaling between the mapping time and the dataset chunk size~$D$:~$T_{mapping}\propto  D\propto 1/N$, see~\cite{bowtie}, and Figure~\ref{fig:fig1}. Since the ratio~$T_{mapping}/T_{comm}\approx \mbox{const}$, one can conclude that the~$T_{comm}\propto 1/N$, meaning that the more nodes are involved, the faster the communication is in the preprocessing stage.

\textit{HPC results}: The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management} is as follows: the data from the delivery location is being split into chunks in parallel, which are simultaneously being pushed to the local scratches of the nodes, allocated by SLURM  (open circles). One can see two distinct linear stretches, each with a different slope. One stretch is for the range from 4 to about 12 nodes, and the another is from 12 up to 60. The former (horizontal stretch) is explained as for Hadoop~-- the more nodes being involved the faster the chunks are being distributed. The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbit/sec of outcoming traffic} and unable to pass more data in a time unit, while the mapping time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{mapping}\propto D\propto 1/N \rightarrow T_{mapping}/T_{comm}\propto 1/N$, i.e. a linear dependency, which one can observe in the plot. 
The transition area between the two modes~-- saturated and unsaturated~-- has the next origin: each hard disk drive on the local node can write the data at a speed about 100MB/sec\,$\approx$\,1Gbit/sec, i.e.~10 nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch. The nodes are being allocated  on the same rack, which is the default SLURM behaviour.
